Isolator: accurate and consistent analysis of RNA-Seq experiments

### 2015-11-07 paper.tex
Combine some tables into one figure, etc.

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{sectsty}
\usepackage{ctable}
\usepackage{hyperref}
\usepackage[numbers]{natbib}
\usepackage{xr}

\externaldocument{supplement}

@@

\title{Isolator: accurate and consistent analysis of RNA-Seq experiments}
\author{Daniel C. Jones \and Kavitha T. Kuppusamy \and Xinxia Peng \and Michael G. Katze \and Hannele Ruohola-Baker \and Walter L. Ruzzo}
\date{}

@@

We use data generated by the Sequencing Quality Control Consortium (SEQC)
\cite{Consortium:2014kc}, consisting of four reference samples (labeled A, B, C,
and D). The RNA-seq methods were run using approximately 25 million reads per
sample, obtaining estimates that were then compared to qPCR in Tables
\ref{table:taqman} and \ref{table:primepcr}, and the known ERCC spike-in
mixtures in Table \ref{table:ercc}. In each of these we find that Isolator
produces the highest correlation of the methods compared, though sometimes by a
negligible margin. Despite two different technologies, and two different sets of
genes, the qPCR benchmarks agree closely, ranking the methods nearly
identically. Cufflinks, eXpress, Salmon, and RSEM/ML typically perform similarly
to each other, ith BitSeq, Kallisto, Sailfish, and RSEM/PM trailing slightly.
Results are similar with ERCC spike-in controls, the notable exceptions being
Kallisto and Sailfish, which show higher accuracy, and eXpress, which shows
lower accuracy compared to their relative rankings in the qPCr benchmarks.

\begin{table}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.903} & \textbf{0.908} & \textbf{0.890} &
\textbf{0.883}
\\\addlinespace
eXpress & 0.901 & 0.902 & 0.884 & 0.874
\\\addlinespace
Cufflinks & 0.900 & 0.901 & 0.865 & 0.872
\\\addlinespace
Salmon & 0.899 & 0.897 & 0.882 & 0.871
\\\addlinespace
RSEM/ML & 0.899 & 0.896 & 0.881 & 0.870
\\\addlinespace
Kallisto & 0.893 & 0.885 & 0.874 & 0.860
\\\addlinespace
BitSeq & 0.875 & 0.865 & 0.863 & 0.848
\\\addlinespace
Sailfish & 0.879 & 0.857 & 0.854 & 0.834
\\\addlinespace
RSEM/PM & 0.878 & 0.866 & 0.864 & 0.848
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 806 genes using TaqMan qPCR and RNA-Seq quantification.}
\label{table:taqman}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.878} & \textbf{0.866} & \textbf{0.839} &
\textbf{0.851}
\\\addlinespace
Cufflinks & 0.870 & 0.856 & 0.799 & 0.841
\\\addlinespace
eXpress & 0.870 & 0.855 & 0.829 & 0.840
\\\addlinespace
Salmon & 0.866 & 0.852 & 0.826 & 0.836
\\\addlinespace
RSEM/ML & 0.865 & 0.851 & 0.825 & 0.835
\\\addlinespace
BitSeq & 0.840 & 0.821 & 0.802 & 0.813
\\\addlinespace
Kallisto & 0.858 & 0.840 & 0.817 & 0.826
\\\addlinespace
Sailfish & 0.844 & 0.814 & 0.797 & 0.802
\\\addlinespace
RSEM/PM & 0.840 & 0.822 & 0.803 & 0.811
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 18,353 genes using PrimePCR qPCR and RNA-Seq quantification.}
\label{table:primepcr}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.980} & \textbf{0.979} & \textbf{0.981} &
\textbf{0.983}
\\\addlinespace
Salmon & 0.976 & 0.975 & 0.978 & 0.979
\\\addlinespace
Kallisto & 0.972 & 0.972 & 0.973 & 0.976
\\\addlinespace
Sailfish & 0.970 & 0.969 & 0.969 & 0.972
\\\addlinespace
Cufflinks & 0.967 & 0.969 & 0.972 & 0.974
\\\addlinespace
RSEM/PM & 0.943 & 0.949 & 0.944 & 0.949
\\\addlinespace
RSEM/ML & 0.941 & 0.948 & 0.945 & 0.951
\\\addlinespace
BitSeq & 0.940 & 0.949 & 0.943 & 0.946
\\\addlinespace
eXpress & 0.931 & 0.939 & 0.935 & 0.942
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between known proportions of 92
ERCC spike-in controls and RNA-Seq quantification.}
\label{table:ercc}
\end{table}

@@

The four SEQC samples consist of two commercially available reference RNA
samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
while sample $D$ is 25\% $A$ and 75\% $B$. This design allows us to directly
measure expression consistency. If we estimate transcript abundance for each
sample, producing estimates $a, b, c, d$, then for these estimates to be in
agreement we should expect $c$ to be approximated by $0.25a + 0.75b$, and
similarly $d$ by $0.75a + 0.25b$. To capture the notion of consistency, we
measured the closeness of this approximation using proportionality correlation
(Table \ref{table:seqctranscripts}).

@@

\begin{table}\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.978} & \textbf{0.978}
\\\addlinespace
BitSeq & 0.967 & 0.967
\\\addlinespace
RSEM/PM & 0.968 & 0.967
\\\addlinespace
Sailfish & 0.932 & 0.925
\\\addlinespace
RSEM/ML & 0.922 & 0.919
\\\addlinespace
Salmon & 0.916 & 0.914
\\\addlinespace
Kallisto & 0.907 & 0.902
\\\addlinespace
eXpress & 0.903 & 0.899
\\\addlinespace
Cufflinks & 0.870 & 0.916
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between transcript-level estimates
for the mixed SEQC samples C and D and weighted averages of estimates for A
and B, corresponding to the mixture proportions for C and D. Sample C consists of
$3/4$ A and $1/4$ B and sample D is $1/4$ A and $3/4$ B. Consistent estimates of
transcript expression in A, B, C, and D should agree with these mixture
proportions.}
\label{table:seqctranscripts}
\end{table}

@@

\begin{table}\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator & 0.919
\\\addlinespace
Kallisto & 0.887
\\\addlinespace
Salmon & 0.886
\\\addlinespace
RSEM/ML & 0.881
\\\addlinespace
Cufflinks & 0.881
\\\addlinespace
eXpress & 0.825
\\\addlinespace
Sailfish & 0.816
\\\addlinespace
RSEM/PM & 0.806
\\\addlinespace
BitSeq & 0.796
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between ground truth and estimates
produced by each method using simulated RNA-Seq reads.}
\label{table:rlsim}
\end{table}

@@

\subsection{Finding monotonic differential splicing}\label{finding-monotonic-differential-splicing}

@@

>>>>

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{sectsty}
\usepackage{ctable}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage[numbers]{natbib}
\usepackage{xr}

% TODO: wtf does this break things?
%\externaldocument{supplement}

@@

\title{Isolator: accurate and consistent analysis of RNA-Seq experiments}

% Possible alternative titles? Larry suggests putting 'isoform' in the name.
% 
% accurate and consistent isoform quantification
%
%
%

\author{Daniel C. Jones \and Kavitha T. Kuppusamy \and Xinxia Peng \and Michael G. Katze \and Hannele Ruohola-Baker \and Walter L. Ruzzo}
\date{}

@@

We use data generated by the Sequencing Quality Control Consortium (SEQC)
\cite{Consortium:2014kc}, consisting of four reference samples (labeled A, B, C,
and D). The RNA-seq methods were run using approximately 25 million reads per
sample, obtaining estimates that were then compared to qPCR in Figures
\ref{table:taqman} and \ref{table:primepcr}, and the known ERCC spike-in
mixtures in Figure \ref{table:ercc}. In each of these we find that Isolator
produces the highest correlation of the methods compared, though sometimes by a
negligible margin. Despite two different technologies, and two different sets of
genes, the qPCR benchmarks agree closely, ranking the methods nearly
identically. Cufflinks, eXpress, Salmon, and RSEM/ML typically perform similarly
to each other, ith BitSeq, Kallisto, Sailfish, and RSEM/PM trailing slightly.
Results are similar with ERCC spike-in controls, the notable exceptions being
Kallisto and Sailfish, which show higher accuracy, and eXpress, which shows
lower accuracy compared to their relative rankings in the qPCr benchmarks.

\begin{figure*}
\begin{subtable}{0.5\linewidth}
\caption{Comparison of estimates to TaqMan qPCR}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.903} & \textbf{0.908} & \textbf{0.890} &
\textbf{0.883}
\\\addlinespace
eXpress & 0.901 & 0.902 & 0.884 & 0.874
\\\addlinespace
Cufflinks & 0.900 & 0.901 & 0.865 & 0.872
\\\addlinespace
Salmon & 0.899 & 0.897 & 0.882 & 0.871
\\\addlinespace
RSEM/ML & 0.899 & 0.896 & 0.881 & 0.870
\\\addlinespace
Kallisto & 0.893 & 0.885 & 0.874 & 0.860
\\\addlinespace
BitSeq & 0.875 & 0.865 & 0.863 & 0.848
\\\addlinespace
Sailfish & 0.879 & 0.857 & 0.854 & 0.834
\\\addlinespace
RSEM/PM & 0.878 & 0.866 & 0.864 & 0.848
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\label{table:taqman}
\end{subtable}
\begin{subtable}{0.5\linewidth}
\caption{Comparison of estimates to PrimePCR qPCR}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.878} & \textbf{0.866} & \textbf{0.839} &
\textbf{0.851}
\\\addlinespace
Cufflinks & 0.870 & 0.856 & 0.799 & 0.841
\\\addlinespace
eXpress & 0.870 & 0.855 & 0.829 & 0.840
\\\addlinespace
Salmon & 0.866 & 0.852 & 0.826 & 0.836
\\\addlinespace
RSEM/ML & 0.865 & 0.851 & 0.825 & 0.835
\\\addlinespace
BitSeq & 0.840 & 0.821 & 0.802 & 0.813
\\\addlinespace
Kallisto & 0.858 & 0.840 & 0.817 & 0.826
\\\addlinespace
Sailfish & 0.844 & 0.814 & 0.797 & 0.802
\\\addlinespace
RSEM/PM & 0.840 & 0.822 & 0.803 & 0.811
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\label{table:primepcr}
\end{subtable}

\begin{subtable}{0.5\linewidth}
\caption{Comparison of estimates to ERCC spike-in controls}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.980} & \textbf{0.979} & \textbf{0.981} &
\textbf{0.983}
\\\addlinespace
Salmon & 0.976 & 0.975 & 0.978 & 0.979
\\\addlinespace
Kallisto & 0.972 & 0.972 & 0.973 & 0.976
\\\addlinespace
Sailfish & 0.970 & 0.969 & 0.969 & 0.972
\\\addlinespace
Cufflinks & 0.967 & 0.969 & 0.972 & 0.974
\\\addlinespace
RSEM/PM & 0.943 & 0.949 & 0.944 & 0.949
\\\addlinespace
RSEM/ML & 0.941 & 0.948 & 0.945 & 0.951
\\\addlinespace
BitSeq & 0.940 & 0.949 & 0.943 & 0.946
\\\addlinespace
eXpress & 0.931 & 0.939 & 0.935 & 0.942
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\label{table:ercc}
\end{subtable}
\begin{subtable}{0.5\linewidth}
\caption{Consistency of SEQC sample estimates}
\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.978} & \textbf{0.978}
\\\addlinespace
BitSeq & 0.967 & 0.967
\\\addlinespace
RSEM/PM & 0.968 & 0.967
\\\addlinespace
Sailfish & 0.932 & 0.925
\\\addlinespace
RSEM/ML & 0.922 & 0.919
\\\addlinespace
Salmon & 0.916 & 0.914
\\\addlinespace
Kallisto & 0.907 & 0.902
\\\addlinespace
eXpress & 0.903 & 0.899
\\\addlinespace
Cufflinks & 0.870 & 0.916
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\label{table:seqctranscripts}
\end{subtable}

\caption{
Benchmarks measuring the accuracy and consistency of estimates made from 
Sequencing Quality Control (SEQC) dataset using a variety of methods.
All numbers are reported in proportionality correlation.
(a) correlation between gene level quantification
of 806 genes using TaqMan qPCR and RNA-Seq quantification.
(b) correlation between gene level quantification
of 18,353 genes using PrimePCR qPCR and RNA-Seq quantification.
(c) correlation between known proportions of 92
ERCC spike-in controls and RNA-Seq quantification.
(d) correlation between transcript-level estimates
for the mixed SEQC samples C and D and weighted averages of estimates for A
and B, corresponding to the mixture proportions for C and D. Sample C consists of
$3/4$ A and $1/4$ B and sample D is $1/4$ A and $3/4$ B. Consistent estimates of
transcript expression in A, B, C, and D should agree with these mixture
proportions.}
\end{figure*}

@@

The four SEQC samples consist of two commercially available reference RNA
samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
while sample $D$ is 25\% $A$ and 75\% $B$. This design allows us to directly
measure expression consistency. If we estimate transcript abundance for each
sample, producing estimates $a, b, c, d$, then for these estimates to be in
agreement we should expect $c$ to be approximated by $0.25a + 0.75b$, and
similarly $d$ by $0.75a + 0.25b$. To capture the notion of consistency, we
measured the closeness of this approximation using proportionality correlation
(Figure \ref{table:seqctranscripts}).

@@

@@

\begin{table}
\label{table:rlsim}
\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
\textbf{Isolator} & \textbf{0.919}
\\\addlinespace
Kallisto & 0.887
\\\addlinespace
Salmon & 0.886
\\\addlinespace
RSEM/ML & 0.881
\\\addlinespace
Cufflinks & 0.881
\\\addlinespace
eXpress & 0.825
\\\addlinespace
Sailfish & 0.816
\\\addlinespace
RSEM/PM & 0.806
\\\addlinespace
BitSeq & 0.796
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{
Proportionality correlation between ground truth and estimates
produced by each method using simulated RNA-Seq reads.}
\end{table}

@@

\subsection{Case study: finding monotonic differential splicing}\label{finding-monotonic-differential-splicing}

@@

\subsection{Case study: novel mechanisms of cell fate specification identified
by gene splice variants}

TODO: add nathan's text here

### 2015-11-07 supplement.tex
Combine some tables into one figure, etc.

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{sectsty}
\usepackage{ctable}
\usepackage{hyperref}
\usepackage[numbers]{natbib}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{analysis/cardio/ACOT9-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/IMMT-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/LTBP4-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/OBSL1-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/OSBPL1A-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/PALM-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/TNNT2-comparison.pdf}
\caption{Point estimates of fold change of exon expression from qPCR and and
RNA-Seq using various methods. Exon expression for RNA-Seq is computed by
estimating transcript expression, then summing the expression of all transcripts
that share a parcitular exon.}
\label{fig:cardiofoldchange}
\end{figure}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{analysis/cardio/isolator-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/salmon-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/sailfish-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/kallisto-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/express-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/cufflinks-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/pm-rsem-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/ml-rsem-pca.pdf}
\caption{The first two principal components from principal component analysis
applied to transcript expression estimates using a variety of methods.}
\label{fig:cardiopca}
\end{figure}

>>>>

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{sectsty}
\usepackage{ctable}
\usepackage{hyperref}
\usepackage[numbers]{natbib}
\usepackage{subcaption}

@@

\begin{figure}[H]
\includegraphics[width=0.2\textwidth]{analysis/cardio/colorkey.pdf} \\
\includegraphics[width=\textwidth]{analysis/cardio/ACOT9-comparison.pdf}
\includegraphics[width=\textwidth]{analysis/cardio/IMMT-comparison.pdf}
\includegraphics[width=\textwidth]{analysis/cardio/LTBP4-comparison.pdf}
\end{figure}
\begin{figure}[H]
\ContinuedFloat
\includegraphics[width=\textwidth]{analysis/cardio/OBSL1-comparison.pdf}
\includegraphics[width=\textwidth]{analysis/cardio/OSBPL1A-comparison.pdf}
\includegraphics[width=\textwidth]{analysis/cardio/PALM-comparison.pdf}
\includegraphics[width=\textwidth]{analysis/cardio/TNNT2-comparison.pdf}
\caption{Point estimates of fold change of exon expression from qPCR and and
RNA-Seq using various methods. Exon expression for RNA-Seq is computed by
estimating transcript expression, then summing the expression of all transcripts
that share a parcitular exon.}
\label{fig:cardiofoldchange}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=0.2\textwidth]{analysis/cardio/colorkey.pdf} \\
\includegraphics[width=0.5\textwidth]{analysis/cardio/isolator-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/salmon-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/sailfish-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/kallisto-pca.pdf}
\end{figure}
\begin{figure}
\ContinuedFloat
\includegraphics[width=0.5\textwidth]{analysis/cardio/express-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/cufflinks-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/pm-rsem-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/ml-rsem-pca.pdf}
\caption{The first two principal components from principal component analysis
applied to transcript expression estimates using a variety of methods.}
\label{fig:cardiopca}
\end{figure}
